---
title: "Analyzing CMPD Traffic Stops"
author: "Augustine Kwabena Osei-Fosu"
date: "03/11/2019"
---

# c

```{r load-packages-data, warning=FALSE, message=FALSE}
options(repos = c(CRAN = "http://cran.rstudio.com"))
library(tidyverse)
library(scales)
library(ggspatial) # make sure to install if you don't have it

df <- read_csv("./data/Officer_Traffic_Stops.csv")
```

## Demographics of drivers (1 point)

First, look at the data using the `glimpse()` function from `dplyr`

```{r glimpse}
glimpse(df)
```

Notice the different variable types: character (chr), num (numeric), and datetime (POSIXct).

Let's consider our target variable: `Was_a_Search_Conducted`.

**Plot** a bar chart that counts the number of records by `Was_a_Search_Conducted`.

```{r bar}
ggplot(df, aes(Was_a_Search_Conducted)) +
  geom_bar(color = "blue")
```

**It's okay to use `geom_col()` instead of `geom_bar()`**
Certainly, it is not okay for one to use geom_col in this situation because that option requires both x and y aesthetics to be able to use it and in this case, we are only given the x aesthetic which is, Was_a_Search_Conducted

How well balanced is the data set by this field?
As far as the field  Was_a_Search_Conducted is concerned dataset is unbalanced since a dataset is said to be unbalanced when the class of interest (Was_a_Search_Conducted (Yes)) is much rarer than normal behavior (Was_a_Search_Conducted (No)).

Next, let's consider the age range of the driver. 

**Plot** a histogram of `Driver_Age`. Determine an appropriate number of bins.

```{r histogram}
ggplot(df, aes(Driver_Age)) +
  geom_histogram(bins = 25)
```

Once you go above (around) 40-50 bins, you'll notice some points stick out. 

What is happening? xxxxx
Because the bin width is too large and the smaller features in the distribution of the data, such as the dips around these peaks disappear making these bigger features peak out since their out of bin distance.

**Plot** a density plot of `Driver_Age`. Add in a fill to be "lightblue". Determine an appropriate kernel density to use (`adjust`).

```{r density}
ggplot(df, aes(Driver_Age)) +
  geom_density(bw = "nrd0", adjust = 1.6, fill="lightblue", kernel = "gaussian")

```

**Plot** a box plot with `Was_a_Search_Conducted` on the x-axis and `Driver_Age` on the y-axis.

```{r boxplot}
ggplot(df, aes(Was_a_Search_Conducted, Driver_Age)) +
  geom_boxplot()

```

**Plot** a violin plot.

```{r violin}
ggplot(df, aes(Was_a_Search_Conducted, Driver_Age)) +
  geom_violin()
```

From the plots above, do you think the age of the driver is a significant factor in whether a search was conducted? Why or why not? 
It is very difficult to tell, whether the age of the person is a factor or not. Since those who got searched seems to fall within the age limit of those who were not searched. I will say it is a factor since the 25th to 75th quartiles of those who got searched are slightly above the median of those who were not searched. Also, their outlier almost falls within the max of those who were not searched.

## Date of stop (1 point)

Let's plot the number of stops by time. 

Recalling part one, the `Month_of_Stop` variable is a character, not a date variable. The datatime's are simply when the data was collected; not when the stop occurred. Therefore, we'll need to convert the `Month_of_Stop` variable from a character to a Date format.

Let's first cleanup the date field using `tidyverse` packages like `stringr` and `lubridate`.

```{r date-cleanup}
library(stringr); library(lubridate)
df <- df %>%
  mutate(Month_of_Stop = str_replace_all(Month_of_Stop, "/","-")) %>% # replace "/" with "-"
  mutate(Month_of_Stop = paste0(df$Month_of_Stop,"-01")) %>% # add in day
  mutate(Date = ymd(Month_of_Stop)) # created a date field
```

**Plot** a line chart with the number of traffic stops for each month (hint: start with the `count()` function by Date then feed into ggplot. Remember the count variable is named 'n'.).

```{r stops}
count(df, "n", Date)%>%
  ggplot(aes(Date, n)) + 
  geom_line()
```

What is the trend (i.e., long term rate of change) of the number of traffic stops in Charlotte? 
It does not appear that there is any clear trend to Charlotte number of traffic stop 

**Plot** the same plot but add in `facet_wrap()` by the `Reason_for_Stop` variable.

```{r stops-by-reason}
count(df, "n", Date, Reason_for_Stop)%>%
    ggplot(aes(Date, n)) + 
    geom_line(aes( color = Reason_for_Stop)) +
    facet_wrap(~Reason_for_Stop) +  theme(legend.position = "none")

```

What is a problem with this plot? xxxxx
The problem with this plot is that some of the reasons for stopping such as "Investigation", "CheckPoint" counts were very low while others such as "Speeding", "Vehicle Regulatory"  were very high hence those having high counts seems to limit the visibility of the low counts since all the charts are using the same scale. We can solve this problem by letting scales to vary across the panels

To address this problem, will need to figure out how to adjust the scale. To do this, need to use R's documentation to see whether there is a parameter in `facet_wrap`.

Will go to RStudio console and type `?facet_wrap`.

What parameter allows you to modify the scales of `facet_wrap`? xxxxx
scales to vary across the panels with the `scales` argument. Thus facet_wrap(~Reason_for_Stop, scales =  "free")

**Plot** the same plot but with a free y-axis scale.

```{r stops-by-reason-2}
count(df, "n", Date, Reason_for_Stop)%>%
    ggplot(aes(Date, n)) + 
    geom_line(aes( color = Reason_for_Stop)) +
    facet_wrap(~Reason_for_Stop, scales =  "free") +  theme(legend.position = "none")
```

Which type of police stop has had the most volatility (i.e., big swings in number of stops)? xxxxx
Speeding

What is one problem with allowing the y-axis be free? xxxxx
If conscious effort n attention is not drawn to the different scales there is a high propensity that people will misinterpret these faceted charts in comparison and also as you rightly stated small multiples tends to be less effective when each of the variables are on different scales or magnitudes.

Let's consider instead CMPD traffic stops but by CMPD division. These are more even spread by division than the type of stop.

**Plot** a line chart (optional points too) for stops by `Date` (x axis) and counts ('n', or whatever you named your count variable) (y axis). (hint: to modify how the date is shown, use the layer `scale_x_date(date_labels = "%Y") +` to show only the year. Feel free to modify by looking at `?scale_x_date`.)

```{r stops-by-division}
count(df, "n", Date, CMPD_Division)%>%
    ggplot(aes(Date, n)) + 
    geom_line(aes( color = CMPD_Division)) + scale_x_date(date_labels = "%Y") +
    facet_wrap(~CMPD_Division, scales = "free") + theme(legend.position = "none")
```

What are three observations you can make about the number of police stops by divison? (hint: just write about what's in the data.)

1. xxxxx
Almost every division' stopping shows a very high volatility 

2. xxxxx
After 2017 almost every division's stopping seems to show some bit of downwards trend 
3. xxxxx
Of all the divisions' Freedom division seems to depict the lowest traffic stop 

Next, this doesn't help tell us where these areas are. For that, let's use a shape file to create a chloropleth of stops by division.

## Geography

For this example, we'll create a cholorpleth for the number of police stops by police division.

To do this, we need to use the `sf` package. (For help along the way, see [this tutorial](http://strimas.com/r/tidy-sf/) on `sf` package.)

```{r load-sf}
library(sf); library(viridis)
cmpd <- st_read("./data/CMPD_Police_Divisions.shp")
```

Note that while we have five files, we only load in the shapefile (`.shp`) file. This is typical but know that to use this file you would need the other four files in the same folder as your shapefile.

**Plot** cmpd using the `geom_sf` package where you provide `fill = DNAME` as the only aesthetic. Add in a title saying "CMPD Divisions" and add the `theme_bw()` theme to make translate the file into the black and white template.

```{r geom_sf}
ggplot(cmpd) +
    geom_sf(aes(fill = DNAME)) +
    ggtitle("CMPD Divisions") +
    theme_bw()
```

One problem with this map is it's hard to read the division names. That is, it may be better to remove the legend and put the labels of each division within the plot.

To do this, we can use the related `geom_sf_label()` geom, using the name of the division as the aesthetic label.

**Plot** the same plot from above but with the name of the division as the label. 

1. You'll likely need to reduce the size of the label, using the `size` paramater. You should likely set the `size` to under 2. 

2. Make sure to remove the legend (it's redundant and no longer necessary).

3. Create a new variable named `Name` that removes the term " Division". This term is redundant and takes up a lot of space in the labels from `DNAME`. To do this step, use this snippet of code at the top of your pipeline:

```{r eval=F}
cmpdN <- cmpd %>%
  mutate(Name = as.character(DNAME)) %>%
  mutate(Name = str_replace_all(Name, " Division",""))
```

4. Save it as an object named `g`. Make sure to call it once so that the map will output.

```{r geom-labels}
g <- ggplot(cmpdN) +
    geom_sf(aes(fill = Name), show.legend = F) + 
  geom_sf_label(aes(label = Name), size = 3) + 
  ggtitle("CMPD Divisions") + theme_bw()
```

**Plot**: Now let's change the projection type. Simply add in `coord_sf(crs = st_crs(102003))` as an additional layer to `g`.

```{r albers}
g + coord_sf(crs = st_crs(102003))
```


Do you notice any difference between the two maps? xxxxx
Yes, It alters the coordinates a little bit in an anticlockwise direction. 

## Advanced Plot:

Now, let's create a chloropleth. Below is the code to create an advanced plot.

In this problem, you need to explain what each step below is doing:

- `mutate()`: 
xxxxx It is creating new variables by name, CMPD_Division  and the method is by forcing the already existing variable, DNAME to a character 
- `inner_join()`: 
xxxxx
It is joining on "CMPD_Division" the counted variables CMPD_Division weight by Date from data frame df.
- `mutate()`: 
xxxxx
It is creating new variables by name, Year and the method is by using year function in lubridate package to assign the year part of Date variable to the new variable.
- `geom_sf()`: 
xxxxx
It is explicitly passing the sf object, n to the data argument of the ggplot to develop the plot
- `scale_fill_viridis()`: 
xxxxx
It is filling the color scale option argument in the ggplot scales of the ggplot developed plot 
- `labs()`: 
xxxxx
It is used to supply the labeling argument of the plot 
- `annotation_scale()`: 
xxxxx
It is adding a scale bar to the plot
- `facet_wrap()`: 
xxxxx
Separating or splitting the created plot by year variable, thus to have each year's plot 
- `theme_bw()`: 
xxxxx
 To make translate the file into the black and white template
- `theme()`: (what are each of the options doing in `theme()`?) 
xxxxx
These are manipulating the labeling and look aspect of the plot, with legend.position pointing to where the legend should be placed on the plot and in this plot it pointed the bottom; plot.title is about how the tile should look; axis.text.x is about how the x axis to look  and while the y counterpart is about the y axis 
- `ggsave()`: 
xxxxx
It is a function which is conveniently used to save the created plot 

```{r adv-plot}
cmpd_chloropleth <- cmpd %>% 
  mutate(CMPD_Division = as.character(DNAME)) %>% 
  inner_join(count(df, CMPD_Division, Date), by = "CMPD_Division") %>%
  mutate(Year = lubridate::year(Date)) %>%
  ggplot() +
  geom_sf(aes(fill = n)) +
  scale_fill_viridis("Traffic Stops", labels = scales::comma) +
  labs(title = "CMPD Traffic stops by CMPD Division",
       caption = "Source: CMPD") +
  annotation_scale(location = "bl", width_hint = 0.2) +
  facet_wrap(~Year) +  
  theme_bw() +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = rel(1.5)),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank())

cmpd_chloropleth

ggsave(cmpd_chloropleth, filename = "cmpd_chloropleth.pdf",
       width = 7, height = 5, units = "in")
ggsave(cmpd_chloropleth, filename = "cmpd_chloropleth.png",
       width = 7, height = 5, units = "in")
```


## Choosing a ggplot extensions

Go to [ggextensions website](https://www.ggplot2-exts.org/). Then click Galleries to explore the different ggplot extensions. Scroll through and see if any catch your eye.

Now, select **one** of the ggextension libraries below and install the package (through CRAN):

- [`ggridges`](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) / [example to recreate](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html#varying-fill-colors-along-the-x-axis)^[Run both plots. Make sure to install the `viridis` package or else you'll get an error!]

- [`ggalt`](https://github.com/hrbrmstr/ggalt) / [example to recreate](https://github.com/hrbrmstr/ggalt#horzon-chart)^[Make sure to install `hrbrthemes`!]

- [`ggrepel`](https://github.com/slowkow/ggrepel) / [example to recreate](https://github.com/slowkow/ggrepel/blob/master/vignettes/ggrepel.md#align-labels-on-the-left-or-right-edge)

- [`ggstatsplot`](https://github.com/IndrajeetPatil/ggstatsplot) / [example to recreate](https://github.com/IndrajeetPatil/ggstatsplot#ggscatterstats)^[Run all three examples in the `ggscatterstats` section.]

**Plot** the related example

```{r ggextension-examples}
library(ggridges); library(ggjoy)
head(lincoln_weather)
library(tidyverse); library(viridisLite) #will be needed if not already loaded
ggplot(lincoln_weather, aes(x = `Mean Temperature [F]`, y = `Month`, fill = ..x..)) +
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis(name = "Temp. [F]", option = "C") +
    labs(title = 'Temperatures in Lincoln NE in 2016')
```

Now, with the same package you ran, make a **plot** with that package and the gapminder data. You can choose any of the data frames (i.e., years). Make sure your plot has at least six functions (e.g., ggplot() + geom_point() is two functions and `dplyr` functions count as well.)

```{r ggextension-custom}
library(ggridges); library(ggjoy) #will be needed if not already loaded
library(tidyverse); library(viridisLite) #will be needed if not already loaded
library(gapminder); library(ggthemes)
ggplot(gapminder, aes( x = lifeExp, 
                       y = continent, fill = continent)) +
geom_density_ridges(alpha = 0.4) + 
  scale_fill_viridis(discrete = T, option = "B", 
                     direction = -1, begin = .1, end = .9)+
    theme_tufte() + 
  theme_minimal(base_family =  "mono") + 
  labs(x = "Life Expectancy",y = "Continent",
       title = "Global convergence in Continental life expectancy for every five years, from 1952 to 2007", subtitle = "An excerpt at Gapminder.org. For each of 142 countries",
       caption = "OFAK") +
  theme(legend.position = "none")
```

xxxxx
I have found that there is so much one can do with ggplot compliment with a lot of other r packages but it seems very impossible to make any decent plot without incorporating tidyverse. Again I can attest to the fact I have come to appreciate the fact that the same date can be visualized in several ways. But you need to teach us how these datasets are organized especially the CMPD data we used in this assignment.

## Intro to interactivity

For even more fun, **plot** an interactive HTML plot using the code for any of the plots above (fair warning, some of the ggextensions may not work well). 

The easiest way to do this is to use the `plotly` package (install it with the "Packages" panel in RStudio), and then to use its `ggplotly()` function.

I've given you some commented-out code below (commented out so that R doesn't yell at you about the code not working when you knit). 

Also, check out [the documentation](https://plot.ly/ggplot2/), especially [this page](https://plot.ly/ggplot2/interactive-tooltip/) about customizing the tooltips that show up when you hover over points or areas.

```{r interactive, message=FALSE})
library(plotly)
library(xfun)
 p = count(df, "n", Date)%>%
   ggplot(aes(Date, n)) + 
   geom_line()
ggplotly(p)
```
